#Assignment 2

---
author: "Ana Zhibaj"
date: "9/11/2020"
output: html_document
---

##Setting up the workspace

I initially installed tidyverse and ggplot2 libraries.

```{r}
library(tidyverse)
library(ggplot2)
library(viridis)
library(ggthemes)
library(wesanderson)
```

I loaded the data I created for the last assignment, that aim to explore the relationship between citizenship, class and race. 

```{r}
citizenship_data <- read_csv("Citizenship-class-race.csv")
attach(citizenship_data)
```

Fret not if you can't remember the variables. That's why I formatted them on this nice table:  

``` {r variable list, echo=FALSE} 
variables_list <- tibble(Name = c("Age", 
                           "Citizenship Status",
                           "Race",
                           "Employment",
                           "Education",
                           "Income",
                           "Year of Entry",
                           "Year of naturalization"),
                         Code = c("AGEP",
                           "CIT",
                           "RACEAIN, RACASN, RACBLK, RACSOR, RACWHT",
                           "ESR",
                           "SCHL",
                           "WAGP",
                           "YOEP",
                           "CITWP"),
                         Type = c("cont",
                                  "cat",
                                  "cat",
                                  "cat",
                                  "cat",
                                  "cont",
                                  "cont",
                                  "cont"))
knitr::kable(variables_list)
``` 

##Visualizing the distribution of continuous variables

Next, I will use the summary function for the continuous variables: Age, Income, Year of Entry, and Period of Naturalization. I will use t.test to get a 95% confidence interval for the population mean. I will use the color palette from Agnes Varda's "Uncle Yanco", a short documentary she directed in 1967 that depicts the life of a distant relative. 

```{r}
age_summary <- summary(AGEP)
age_sd <- sd(AGEP, na.rm = TRUE)
age_pop_mean_int <- t.test(AGEP)
income_summary <- summary(WAGP)
income_sd <- sd(WAGP, na.rm = TRUE)
income_pop_mean_int <- t.test(WAGP)
year_entry_summary <- summary(YOEP)
year_entry_sd <- sd(YOEP, na.rm = TRUE)
year_entry_pop_mean_int <- t.test(YOEP)
periodnat_summary <- summary(years_to_nat)
periodnat_sd <- sd(years_to_nat, na.rm = TRUE)
periodnat_pop_mean_int <- t.test(years_to_nat)
```

I'll format my findings in a pleasant table. 

```{r}
summary_table <- tibble(Statistic = c("Sample mean", 
                "Median",
                "Standard deviation",
                "Interquartile range",
                "Population mean (95% confidence)"),
  `Median age (years)` = c(prettyNum(age_summary["Mean"], digits=3),
                           age_summary["Median"],
                           prettyNum(age_sd, digits = 3),
                           paste(age_summary["1st Qu."],"to", 
                                 age_summary["3rd Qu."]),
                           paste(prettyNum(age_pop_mean_int$conf.int[1],
                                           digits=3), "to",
                                 prettyNum(age_pop_mean_int$conf.int[2],
                                               digits=3))),
    `Income` = c(prettyNum(income_summary["Mean"], digits=3),
                           income_summary["Median"],
                           prettyNum(income_sd, digits = 3),
                           paste(income_summary["1st Qu."],"to", 
                                 income_summary["3rd Qu."]),
                           paste(prettyNum(income_pop_mean_int$conf.int[1],
                                           digits=3), "to",
                                 prettyNum(income_pop_mean_int$conf.int[2],
                                               digits=3))))
knitr::kable(summary_table,
             caption = "Summary")
```

##Histograms

I will use ggplot2 to create a histogram for the distribution. The color selection is inspired by Vardas "Uncle Yanco". 

###Age

```{r}
ggplot(citizenship_data, aes(x = AGEP)) +
  geom_histogram(color = "white", binwidth = 2, fill = "red 2") +
  theme_classic() +
  scale_x_continuous(name = "Age (in years)") +
  scale_y_continuous(name = "Number of people in sample")
```



###Income

```{r}
ggplot(citizenship_data, aes(x = WAGP)) +
  geom_histogram(color = "white", binwidth = 200000, fill = "darkorchid4") +
  theme_classic() +
  scale_x_continuous(name = "Wage") +
  scale_y_continuous(name = "Number of people in sample")
```
###Period of naturalization

```{r}
ggplot(citizenship_data, aes(x = years_to_nat)) +
  geom_histogram(color = "white", binwidth = 1, fill = "gold1") +
  theme_classic() +
  scale_x_continuous(name = "Period of naturalization") +
  scale_y_continuous(name = "Number of people in sample")
```

```{r}
summary(years_to_nat)
```

```{r}
paste("The sample mean for period of naturalization is",
      prettyNum(mean(years_to_nat), digits = 0), "years.")
paste("The sample median for period of naturalization is",
      prettyNum(median(years_to_nat), digits = 0,), "years.")
paste("The sample standard deviation for period of naturalization is",
      prettyNum(sd(years_to_nat), digits = 0), "years")
paste("The sample interquartile range for period of naturalization is",
      quantile(years_to_nat)["25%"], "to", quantile(years_to_nat)["75%"], "years.")
```

What about the t.test? 

```{r}
pop_mean <- t.test(years_to_nat)
pop_mean
```
```{r}
hist <- ggplot(citizenship_data, aes(x = years_to_nat)) + 
  geom_histogram(
    binwidth = 1
  )

hist + 
  theme_classic() +
  scale_x_continuous(name = "Period of citizenship by naturalization (years)") +
  scale_y_continuous(name = "Number of people in sample") +
  geom_histogram(color = "white", binwidth = 1, fill = "navajowhite3") +
  annotate("text", x = 30, y = 1000, size = 3, 
           label = paste("Population mean (95% confidence):",
                         prettyNum(pop_mean$conf.int[1], digits = 3),
                         "to",
                         prettyNum(pop_mean$conf.int[2], digits = 3),
                         "years.",
                         "\nSample mean:", 
                         prettyNum(mean(years_to_nat), digits = 0), "years.",
                         "\nSample median:",
                         prettyNum(median(years_to_nat), digits = 0), "years.",
                         "\nSample standard deviation:",
                         prettyNum(sd(years_to_nat), digits = 0), "years.", 
                         "\nSample interquartile range:",
                         prettyNum(quantile(years_to_nat)["25%"], digits = 0),
                         "to",
                         prettyNum(quantile(years_to_nat)["75%"], digits = 0),
                         "years."),
           hjust = 0, vjust = 1)
```


##Visualizing the distribution of categorical variables


###Employment

Let's calculate the shares for each of the categories of employment. 
```{r}
employment <- unique(ESR_label)
employment
```

```{r}
not_in_labor_force <- t.test(ESR_label == employment [1])
civ_empl_work <- t.test(ESR_label == employment [2])
armed_forces_work <- t.test(ESR_label == employment [3])
unemployed <- t.test(ESR_label == employment [4])
civ_empl_not_work <- t.test(ESR_label == employment [5])

shares_employment <- tibble(Employment = c("Not in labor force", 
                            "Civilian employed, at work", 
                            "Armed forces, at work",
                            "Unemployed",
                            "Civilian employed, but not at work"),
                 `Share` = c(not_in_labor_force$estimate, 
                             civ_empl_work$estimate,
                             armed_forces_work$estimate,
                             unemployed$estimate,
                             civ_empl_not_work$estimate),
                 Low = c(not_in_labor_force$conf.int[1],
                         civ_empl_work$conf.int[1],
                          armed_forces_work$conf.int[1],
                         unemployed$conf.int[1],
                         civ_empl_not_work$conf.int[1]),
                 High = c(not_in_labor_force$conf.int[2],
                         civ_empl_work$conf.int[2],
                          armed_forces_work$conf.int[2],
                         unemployed$conf.int[2],
                         civ_empl_not_work$conf.int[2]))
shares_employment
```

I will visualize these results with a bar chart. 

```{r}
ggplot(shares_employment, aes(x = Employment, y = Share, ymin = Low, ymax = High)) +
  geom_bar(stat = "identity") +
  geom_errorbar(width = 0.5) +
  theme_classic() +
   theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(name = "Share of population (error bar 95-percent confidence interval)",
                     breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
                     labels = c("0", "10%", "20%", "30%", "40%", "50%"))
```

###Level of education

```{r}
education <- unique(SCHL_label)
education
```

```{r}
no_schooling <- t.test(SCHL_label == education [1])
some_college <- t.test(SCHL_label == education [2])
high_school <- t.test(SCHL_label == education [5])
associate_degree <- t.test(SCHL_label == education [6])
college_1 <- t.test(SCHL_label == education [8])
ged <- t.test(SCHL_label == education [9])
masters <- t.test(SCHL_label == education [10])
bachelors <- t.test(SCHL_label == education [15])
prof_degree <- t.test(SCHL_label == education [19])
doctorate <- t.test(SCHL_label == education [20])

shares_education <- tibble(Education = c("No schooling", 
                            "Some college, but less than 1 year",
                            "Regular high school diploma",
                            "Associate's degree",
                            "1 or more years of college credit, no degree",
                            "GED",
                            "Master's degree",
                            "Bachelor's degree",
                            "Professional degree beyond a bachelor's degree",
                            "Doctorate degree"),
                 `Share` = c(no_schooling$estimate, 
                             some_college$estimate,
                             high_school$estimate,
                             associate_degree$estimate,
                             college_1$estimate,
                             ged$estimate,
                             masters$estimate,
                             bachelors$estimate,
                             prof_degree$estimate,
                             doctorate$estimate),
                 Low = c(no_schooling$conf.int[1],
                         some_college$conf.int[1],
                          high_school$conf.int[1],
                         associate_degree$conf.int[1],
                         college_1$conf.int[1],
                         ged$conf.int[1],
                         masters$conf.int[1],
                         bachelors$conf.int[1],
                         prof_degree$conf.int[1],
                         doctorate$conf.int[1]),
                 High = no_schooling$conf.int[2],
                         some_college$conf.int[2],
                          high_school$conf.int[2],
                         associate_degree$conf.int[2],
                         college_1$conf.int[2],
                         ged$conf.int[2],
                         masters$conf.int[2],
                         bachelors$conf.int[2],
                         prof_degree$conf.int[2],
                         doctorate$conf.int[2])
shares_education
```

I will visualize these results with a bar chart. 

```{r}
ggplot(shares_education, aes(x = Education, y = Share, ymin = Low, ymax = High)) +
  geom_bar(stat = "identity") +
  geom_errorbar(width = 0.5) +
  theme_classic() +
  scale_y_continuous(name = "Share of population (error bar 95-percent confidence interval)",
                     breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
                     labels = c("0", "10%", "20%", "30%", "40%", "50%")) +
  coord_flip()
```





