---
title: "Citizenship, class and race"
author: "Ana Zhibaj"
date: "9/8/2020"
output: html_document
---
```{r Load libraries}
library(tidycensus)
library(tidyverse)
library(dplyr)
```

## Dataset variables
I am interested in class and citizenship on an individual level, that is why I'll use variables from the 2018 ACS1 dataset for the state of California. My sample population is citizens that have attained their citizenship by naturalization. I want to inspect how that status (and its period) compares, or corellates, to race and class. 

The variables are as follows:

``` {r variable list, echo=FALSE} 
variables_list <- tibble(Name = c("Age", 
                           "Citizenship Status",
                           "Race",
                           "Employment",
                           "Education",
                           "Income",
                           "Year of Entry",
                           "Year of naturalization"),
                         Code = c("AGEP",
                           "CIT",
                           "RACEAIN, RACASN, RACBLK, RACSOR, RACWHT",
                           "ESR",
                           "SCHL",
                           "WAGP",
                           "YOEP",
                           "CITWP"),
                         Type = c("cont",
                                  "cat",
                                  "cat",
                                  "cat",
                                  "cat",
                                  "cont",
                                  "cont",
                                  "cont"))
knitr::kable(variables_list)
``` 


## Uploading the person-level microdata variables from the 2018 ACS1 census

```{r load person-level microdata variables}
person_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code,
           var_label, data_type, level) %>%
  filter(level == "person", year == 2018, survey == "acs1")
```

## Variable-picking: like apple-picking, only more fun! 
This is where I picked the variables that I had previously defined. I picked multiple indicators for race, because I wanted to have a more nuanced understanding of race, as it compares to citizenship, other than the bland white/ non-white binary. 

```{r, results = 'hide'}
person_population <- get_pums(variables = c("AGEP", 
                                      "CIT", 
                                      "RACAIAN",
                                      "RACASN",
                                      "RACBLK",
                                      "RACSOR", 
                                      "RACWHT",
                                      "ESR",
                                      "SCHL",
                                      "WAGP",
                                      "YOEP",
                                      "CITWP"),
                        state = "CA", 
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  select(AGEP, CIT_label, RACAIAN, RACASN, RACBLK, RACSOR, RACWHT, ESR_label, SCHL_label, WAGP, YOEP, CITWP)
```
I defined the total population by calculating the number of observations from the census. 

```{r Creating the population for the dataset}
total_pop <- nrow(person_population)
```

## Isolating a variable
I filtered the value 4 for Citizenship status, "US Citizen by naturalization", to generate my sample population.

```{r load person-level data, results = 'hide'}
zhibaj_persondata <- get_pums(variables = c( AGEP= "AGEP", 
                                      CIT = "CIT", 
                                      RACAIAN = "RACAIAN",
                                      RACASN = "RACASN",
                                      RACBLK = "RACBLK",
                                      RACSOR = "RACSOR", 
                                      RACWHT = "RACWHT",
                                      ESR = "ESR",
                                      SCHL = "SCHL",
                                      WAGP = "WAGP",
                                      YOEP = "YOEP",
                                      CITWP = "CITWP"),
                        state = "CA", 
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  filter(AGEP > 18) %>%
  filter(CIT == 4) %>%
  mutate(years_to_nat = CITWP - YOEP) %>%
  filter(ESR == 1:6) %>%
  filter(years_to_nat > 0) %>%
  select(AGEP, RACAIAN, RACASN, RACBLK, RACSOR, RACWHT, ESR_label, SCHL_label, WAGP, YOEP, CITWP, years_to_nat)
```
## Sneak peek of the code 

```{r}
zhibaj_persondata
```

## Saving the dataset

```{r}
write.csv(zhibaj_persondata, "Citizenship-class-race.csv")
```



