---
title: "200908_evictions Zhibaj"
author: "Ana Zhibaj"
date: "9/8/2020"
output: html_document
---

```{r, message=FALSE}
library(tidycensus)
library(tidyverse)
```

## Dataset variables
The dataset consists of county-level information on evictions for the state of California in 2015. I combined data from the ACS1 Census of 2015 and eviction data from the California Judicial Council. The variables are as follows:

1. Median Age: Median_age
2. Part of the civilian labor force population that is unemployed: Unemployed_labor
3. Race
4. Period of entry by Nativity: year_entry
5. Reported cases of evictions: State_Reported_Cases

```{r load area-level variables from ACS5 2015}
pop_vars <- c(Median_age =           "B01002_001",
              Unemployed_labor =       "B23025_005",
              total_pop =         "B02001_001", 
              white_alone =       "B03002_003",
              black_AA_alone =    "B03002_004",
              native_alone =      "B03002_005",
              asian_alone =       "B03002_006",
              hawaiian_PI_alone = "B03002_007",
              other =             "B03002_008",
              mult_races =        "B03002_009",
              hispanic_latino =   "B03002_012",
              year_entry =      "B05005_001")
countyCA2015 <- get_acs(geography= "county", 
              variables = pop_vars, 
              output = "wide", 
              state = "CA",
              year = 2015, 
              survey = "acs5")
```

```{r Define countyCA as dataframe}
countyCA <- data.frame(countyCA2015)
```


```{r Load eviction data from the California Judicial Council saved as a CSV}
eviction_dt <- read_csv("calistatsevictions_2.csv")
```

I tidied the CSV by removing columns I wasn't interested in. 

```{r}
eviction_dt <- eviction_dt %>%
  filter(year == 2015)
```

```{r Define Eviction_data as dataframe}
Eviction_data <- data.frame(eviction_dt) 
```


```{r Join dataframes}
Eviction_CA <- inner_join(countyCA, Eviction_data, by = "NAME")
```

```{r}
Eviction_CA
```