---
title: "Assignment 5"
author: "Ana Zhibaj"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, message=FALSE}
library(tidyverse)
library(ggthemes)
library(jtools)
library(interactions)
library(knitr)
```

I changed some of the variables from the previous assignment, namely I reduced the number of categories for Education and Employment status. They offered a level of specificity that did not contribute to my research question. I have filtered the variable for Race, excluding the category "No data", and Employment, excluding the category "Not in labor force".  

```{r load csv, message=FALSE}
citizenship_data <- read_csv("Citizenship-class-race-updated.csv") %>% 
  mutate(SCHL_label = case_when(
    SCHL_label=="1 or more years of college credit, no degree"~"1HS Graduate or lower", # I added 1 in front of the base category. 
    SCHL_label=="12th grade - no diploma"~"1HS Graduate or lower",
    SCHL_label=="Associate's degree"~"Associate's degree",
    SCHL_label=="Bachelor's degree"~"Bachelor's degree or Higher",
    SCHL_label=="Doctorate degree"~"Bachelor's degree or Higher",
    SCHL_label=="GED or alternative credential"~"1HS Graduate or lower",
    SCHL_label=="Grade 10"~"1HS Graduate or lower",
    SCHL_label=="Grade 11"~"1HS Graduate or lower",
    SCHL_label=="Grade 1"~"1HS Graduate or lower",
    SCHL_label=="Grade 2"~"1HS Graduate or lower",
    SCHL_label=="Grade 3"~"1HS Graduate or lower",
    SCHL_label=="Grade 4"~"1HS Graduate or lower",
    SCHL_label=="Grade 5"~"1HS Graduate or lower",
    SCHL_label=="Grade 6"~"1HS Graduate or lower",
    SCHL_label=="Grade 7"~"1HS Graduate or lower",
    SCHL_label=="Grade 8"~"1HS Graduate or lower",
    SCHL_label=="Grade 9"~"1HS Graduate or lower",
    SCHL_label=="Kindergarten"~"1HS Graduate or lower",
    SCHL_label=="Master's degree"~"Bachelor's degree or Higher",
    SCHL_label=="No schooling completed"~"No Schooling Completed",
    SCHL_label=="Nursery school, preschool"~"No Schooling Completed",
    SCHL_label=="Professional degree beyond a bachelor's degree"~"Bachelor's degree or Higher",
    SCHL_label=="Regular high school diploma"~"1HS Graduate or lower",
    SCHL_label=="Some college, but less than 1 year"~"1HS Graduate or lower")) %>%
  mutate(RACE = case_when(
    RACE=="White"~"1White",
    RACE=="Asian"~"Asian",
    RACE=="American Indian and Alaska Native"~"AIAN",
    RACE=="Black"~"Black",
    RACE=="No data"~"No data",
    RACE=="Other"~"Other")) %>%
  mutate(ESR_label = case_when(
    ESR_label=="Unemployed"~"1Unemployed",
    ESR_label=="Armed forces, with a job but not at work"~"Employed",
    ESR_label=="Civilian employed, at work"~"Employed",
    ESR_label=="Civilian employed, with a job but not at work"~"Employed",
    ESR_label=="Not in labor force"~"Not in labor force",
    ESR_label=="Not in labor force "~"Not in labor force ",
    ESR_label=="Armed forces, at work"~"Employed")) %>%
  filter(ESR_label != "N/A (less than 16 years old)") %>%
  filter(RACE != "No data") %>%
  filter(ESR_label != "Not in labor force")
```


## Model 1

This is the intial regression model. The dependent variable is income, and the independent variables are race, employment, education and period of citizenship.

```{r run model 1, message=FALSE}
model1 <- lm(WAGP ~ RACE + ESR_label + SCHL_label + years_to_nat,
            data = citizenship_data)
summary(model1)
```
The adjusted R-squared value is 0.14, meaning this model predicts circa 14% of variation in income.

## Model 2

Adding an interaction term for race and visualizing it, to see if there are differences in income due to race. 

```{r run model 2, message=FALSE}
model2 <- lm(WAGP ~ RACE + ESR_label + SCHL_label + years_to_nat + years_to_nat:RACE + SCHL_label:RACE + ESR_label:RACE,
             data = citizenship_data)
summary(model2)
```

The adjusted R-squared value has increased slightly. We see that there is little to no statistical significance to adding race as an interaction term (we can devise this from the high p-values overall).

## Model 3

Adding an interaction term for employment and visualizing it. 

```{r run model 3, message=FALSE}
model3 <- lm(WAGP ~ RACE + ESR_label + SCHL_label + years_to_nat + years_to_nat:ESR_label + SCHL_label:ESR_label + RACE:ESR_label,
             data = citizenship_data)
summary(model3)
```

## Model 4

Adding an interaction term for education and visualizing it. 

```{r run model 4, message=FALSE}
model4 <- lm(WAGP ~ RACE + ESR_label + SCHL_label + years_to_nat + years_to_nat:SCHL_label + ESR_label:SCHL_label + RACE:SCHL_label,
             data = citizenship_data)
summary(model4)
```

The adjusted R-squared value is 0.15, a 1% improve from the previous models. We can see that there's a statistically significant relationship between having a bachelor's degree or higher, and income. We can observe this from the low p-value:
-People that have a bachelor's degree or higher and period of citizenship is related on average to an income 300 dollars more (p-value of 0.005). 
-People that have a bachelor's degree or higher and are employed have on average an income 44000 dollars more (p-value of 2e-16).
=People belong to other race (not specified on the list) and have a bachelor's degree or higher, make on average 32260 less (p-value of 2e-16).

## Model 5

I will drop the variable for education.

```{r run model 5, message=FALSE}
model5 <- lm(WAGP ~ RACE + ESR_label + years_to_nat,
             data = citizenship_data)
summary(model5)
```
I can see that the value of adjusted R-square dropped to 0.03, so I will try to add an interaction term for employment, although I'm sure the R-squared will be significantly lower. 


## Model 6

I will add an interaction term for employment. 

```{r run model 6, message=FALSE}
model6 <- lm(WAGP ~ RACE + ESR_label + years_to_nat + years_to_nat:ESR_label + RACE:ESR_label,
             data = citizenship_data)
summary(model6)
```

The adjusted R-sqaured in 0.04, a very low value, compared to models 1 through 4.  

## Visualizing the results

I will visualize the results of the adjusted R-squared values for each of the models. 

```{r visualize r squared differences, message=FALSE}
ModelFit <- tibble(model = c(1, 2, 3, 4, 5, 6),
                   R_square = c(summary(model1)$adj.r.squared,
                                summary(model2)$adj.r.squared,
                                summary(model3)$adj.r.squared,
                                summary(model4)$adj.r.squared,
                                summary(model5)$adj.r.squared,
                                summary(model6)$adj.r.squared))

ggplot(ModelFit, aes(x = model, y = R_square)) +
  geom_line() +
  scale_x_continuous(name = "",
                   breaks = breaks <- seq(1, 6, by = 1),
                   labels = paste("Model", breaks)) +
  scale_y_continuous(name = "Adjusted R-squared value") +
  theme_economist()
```

We can see the dip at model 5 and 6. Model 4 is the best fit. On the next step, I will visualize the results on a table. 

## Model results on a table

model1 <- lm(WAGP ~ RACE + ESR_label + SCHL_label + years_to_nat,
            data = citizenship_data)

```{r knit table, message=FALSE}
coeff_names <- c("Race American Indian \n Alaska Native \n(Compared to White)" = "RACEAIAN",
                 "Race Asian \n(Compared to White)" = "RACEAsian",
                 "Race Black \n(Compared to White)" = "RACEBlack",
                 "Race Other \n(Compared to White)" = "RACEOther",
                 "Employed \n(Compared to unemployed)" = "ESR_labelEmployed",
                 "Education \nAssociate's degree \n(Compared to high school degree)" = "SCHL_labelAssociate's degree",
                 "Education \nBachelor's degree or Higher \n(Compared to high school degree)" = "SCHL_labelBachelor's degree or Higher",
                 "Education \nNo Schooling Completed \n(Compared to high school degree)" = "SCHL_labelNo Schooling Completed",
                 "Period of naturalization" = "years_to_nat",
                 "Interaction \nPeriod of naturalization and \nAssociate's degree" = "SCHL_labelAssociate's degree:years_to_nat",
                 "Interaction \nPeriod of naturalization and \nBachelor's degree or Higher" = "SCHL_labelBachelor's degree or Higher:years_to_nat",
                 "Interaction \nPeriod of naturalization and \nNo Schooling Completed" = "SCHL_labelNo Schooling Completed:years_to_nat",
                 "Interaction \nEmployed and \nAssociate's degree" = "ESR_labelEmployed:SCHL_labelAssociate's degree",
                 "Interaction \nEmployed and \nBachelor's degree or Higher" = "ESR_labelEmployed:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nEmployed and \nNo Schooling Completed" = "ESR_labelEmployed:SCHL_labelNo Schooling Completed",
                 "Interaction \nAmerican Indian \n Alaska Native \nand Associate's degree" = "RACEAIAN:SCHL_labelAssociate's degree",
                 "Interaction \nAsian \nand Associate's degree" = "RACEAsian:SCHL_labelAssociate's degree",
                 "Interaction \nBlack \nand Associate's degree" = "RACEBlack:SCHL_labelAssociate's degree",
                 "Interaction \nOther \nand Associate's degree" = "RACEOther:SCHL_labelAssociate's degree",
                 "Interaction \nAmerican Indian \n Alaska Native \nand Bachelor's degree or Higher" = "RACEAIAN:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nAsian \nand Bachelor's degree or Higher" = "RACEAsian:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nBlack \nand Bachelor's degree or Higher" = "RACEBlack:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nOther \nand Bachelor's degree or Higher" = "RACEOther:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nAmerican Indian \n Alaska Native \nand No Schooling Completed" = "RACEAIAN:SCHL_labelNo Schooling Completed",
                 "Interaction \nAsian \nand No Schooling Completed" = "RACEAsian:SCHL_labelNo Schooling Completed",
                 "Interaction \nBlack \nand No Schooling Completed" = "RACEBlack:SCHL_labelNo Schooling Completed",
                 "Interaction \nOther \nand No Schooling Completed" = "RACEOther:SCHL_labelNo Schooling Completed")
export_summs(model1, model4, 
             error_format = "(p = {p.value})",
             error_pos = "same",
             model.names = c("Initial model", "Preferred model"),
             coefs = coeff_names)
```

```{r, model differences, fig.height=15, message=FALSE}
coeff_desc <- c("Race American Indian \n Alaska Native \n(Compared to White)" = "RACEAIAN",
                 "Race Asian \n(Compared to White)" = "RACEAsian",
                 "Race Black \n(Compared to White)" = "RACEBlack",
                 "Race Other \n(Compared to White)" = "RACEOther",
                 "Employed \n(Compared to unemployed)" = "ESR_labelEmployed",
                 "Education \nAssociate's degree \n(Compared to high school degree)" = "SCHL_labelAssociate's degree",
                 "Education \nBachelor's degree or Higher \n(Compared to high school degree)" = "SCHL_labelBachelor's degree or Higher",
                 "Education \nNo Schooling Completed \n(Compared to high school degree)" = "SCHL_labelNo Schooling Completed",
                 "Interaction \nPeriod of naturalization and \nAssociate's degree" = "SCHL_labelAssociate's degree:years_to_nat",
                 "Interaction \nPeriod of naturalization and \nBachelor's degree or Higher" = "SCHL_labelBachelor's degree or Higher:years_to_nat",
                 "Interaction \nPeriod of naturalization and \nNo Schooling Completed" = "SCHL_labelNo Schooling Completed:years_to_nat",
                 "Interaction \nEmployed and \nAssociate's degree" = "ESR_labelEmployed:SCHL_labelAssociate's degree",
                 "Interaction \nEmployed and \nBachelor's degree or Higher" = "ESR_labelEmployed:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nEmployed and \nNo Schooling Completed" = "ESR_labelEmployed:SCHL_labelNo Schooling Completed",
                 "Interaction \nAmerican Indian \n Alaska Native \nand Associate's degree" = "RACEAIAN:SCHL_labelAssociate's degree",
                 "Interaction \nAsian \nand Associate's degree" = "RACEAsian:SCHL_labelAssociate's degree",
                 "Interaction \nBlack \nand Associate's degree" = "RACEBlack:SCHL_labelAssociate's degree",
                 "Interaction \nOther \nand Associate's degree" = "RACEOther:SCHL_labelAssociate's degree",
                 "Interaction \nAmerican Indian \n Alaska Native \nand Bachelor's degree or Higher" = "RACEAIAN:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nAsian \nand Bachelor's degree or Higher" = "RACEAsian:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nBlack \nand Bachelor's degree or Higher" = "RACEBlack:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nOther \nand Bachelor's degree or Higher" = "RACEOther:SCHL_labelBachelor's degree or Higher",
                 "Interaction \nAmerican Indian \n Alaska Native \nand No Schooling Completed" = "RACEAIAN:SCHL_labelNo Schooling Completed",
                 "Interaction \nAsian \nand No Schooling Completed" = "RACEAsian:SCHL_labelNo Schooling Completed",
                 "Interaction \nBlack \nand No Schooling Completed" = "RACEBlack:SCHL_labelNo Schooling Completed",
                 "Interaction \nOther \nand No Schooling Completed" = "RACEOther:SCHL_labelNo Schooling Completed")

plot_coefs(model4, 
           coefs = coeff_desc,
           plot.distributions = TRUE,
           rescale.distributions = TRUE) +
  scale_x_continuous(name = "Effect on income (dollars)")
```

## Visualizing education and income
I will now visualize education and income, by employment, holding all the other variables at their means.

```{r model education and income, message=FALSE}
cat_plot(model3, pred = SCHL_label, modx = ESR_label, interval = TRUE) +
 scale_x_discrete(name = "Education",
                    labels = c("High school degree", "Associate's degree",
                               "Bachelor's degree or Higher", "No schooling completed")) +
  scale_color_discrete(name = "") +
  scale_linetype_discrete(name = "") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_discrete(name = "") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(name = "Income",
                     breaks = seq(-25000, 100000, by = 25000))
```

Next steps: I will try to simplify the categories for education, and run the regression models.


